\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage[backend=biber,style=numeric-comp]{biblatex}
\addbibresource{../../references/references.bib}
\usepackage{hyperref}

\title{Bibliography Example in LaTeX}
\author{Your Name}
\date{\today}

\begin{document}

\maketitle

\section{Introduction}
This document demonstrates how to use bibliography in LaTeX. We'll show different citation styles and bibliography formats.

\section{Citation Examples}
Here are examples of different citation styles:

\subsection{Basic Citations}
According to Smith~\cite{smith2023}, the results show significant improvement.

Recent studies~\cite{jones2022,doe2023} have demonstrated the effectiveness of this approach.

\subsection{Author-Year Citations}
The work by Smith et al.~\cite{smith2023} provides a comprehensive analysis.

As shown in~\cite{jones2022}, the methodology has been validated.

\subsection{Multiple Citations}
Several studies~\cite{smith2023,jones2022,doe2023,brown2021} have investigated this problem.

\section{Literature Review}
The field has evolved significantly over the past decade. Early work by Smith~\cite{smith2023} established the theoretical foundation. Subsequent research by Jones~\cite{jones2022} extended these findings to practical applications.

More recently, Doe~\cite{doe2023} proposed a novel approach that addresses the limitations of previous methods. Brown~\cite{brown2021} provided a comprehensive survey of the field.

\section{Methodology}
Our approach builds upon the work of Smith~\cite{smith2023} and incorporates the improvements suggested by Jones~\cite{jones2022}.

\section{Results}
The results confirm the findings reported in~\cite{doe2023} and extend them to a broader range of applications.

\section{Discussion}
Our findings are consistent with the theoretical predictions of Smith~\cite{smith2023} and provide empirical support for the framework proposed by Jones~\cite{jones2022}.

\section{Conclusion}
This work contributes to the growing body of research in this field, building upon the foundations established by Smith~\cite{smith2023}, Jones~\cite{jones2022}, and Doe~\cite{doe2023}.

% Bibliography
\printbibliography

\end{document}
