\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage[backend=biber,style=numeric-comp]{biblatex}
\addbibresource{../../references/references.bib}
\usepackage{hyperref}

\title{Citation Examples in LaTeX}
\author{Your Name}
\date{\today}

\begin{document}

\maketitle

\section{Basic Citation Styles}
Here are examples of different citation styles:

\subsection{Author-Year Style}
The work by Smith et al.~\cite{smith2023} provides a comprehensive analysis of the problem.

Recent studies~\cite{jones2022,doe2023} have demonstrated the effectiveness of this approach.

\subsection{Numbered Style}
The methodology described in~\cite{smith2023} has been widely adopted.

Several studies~\cite{jones2022,doe2023,brown2021} have investigated this problem.

\subsection{Mixed Citations}
Smith~\cite{smith2023} proposed the initial framework, which was later extended by Jones~\cite{jones2022} and Doe~\cite{doe2023}.

\section{Advanced Citation Techniques}
Here are more advanced citation techniques:

\subsection{Multiple Authors}
The work by Smith, Jones, and Brown~\cite{smith2023} represents a significant contribution.

\subsection{Page References}
As discussed in Smith~\cite[pp. 123-125]{smith2023}, the results are consistent with our findings.

\subsection{Chapter References}
The theoretical foundation is presented in Chapter 3 of Smith~\cite[Chapter 3]{smith2023}.

\section{Citation in Different Contexts}
Here are examples of citations in different contexts:

\subsection{Literature Review}
The field has evolved significantly over the past decade. Early work by Smith~\cite{smith2023} established the theoretical foundation. Subsequent research by Jones~\cite{jones2022} extended these findings to practical applications.

\subsection{Methodology}
Our approach builds upon the work of Smith~\cite{smith2023} and incorporates the improvements suggested by Jones~\cite{jones2022}.

\subsection{Results}
The results confirm the findings reported in~\cite{doe2023} and extend them to a broader range of applications.

\subsection{Discussion}
Our findings are consistent with the theoretical predictions of Smith~\cite{smith2023} and provide empirical support for the framework proposed by Jones~\cite{jones2022}.

\section{Citation Best Practices}
Here are best practices for citations:

\subsection{Appropriate Citation Placement}
\begin{itemize}
    \item Place citations at the end of sentences when possible
    \item Use citations to support claims and statements
    \item Avoid over-citing the same source
    \item Use multiple sources to support important points
\end{itemize}

\subsection{Citation Style Guidelines}
\begin{itemize}
    \item Be consistent with citation style throughout the document
    \item Use appropriate citation format for your field
    \item Include all necessary information in citations
    \item Verify all citations for accuracy
\end{itemize}

\section{Common Citation Mistakes}
Here are common citation mistakes to avoid:

\subsection{Over-citing}
\begin{itemize}
    \item Avoid citing the same source multiple times in one paragraph
    \item Don't cite every sentence
    \item Use citations strategically to support key points
\end{itemize}

\subsection{Under-citing}
\begin{itemize}
    \item Always cite claims that are not common knowledge
    \item Cite all sources of data and statistics
    \item Include citations for methodologies and approaches
\end{itemize}

\section{Citation Management}
Here are tips for managing citations:

\subsection{Organization}
\begin{itemize}
    \item Keep a master bibliography file
    \item Use consistent formatting
    \item Organize citations by topic or type
    \item Use citation management software when possible
\end{itemize}

\subsection{Verification}
\begin{itemize}
    \item Verify all citations for accuracy
    \item Check that all cited sources are accessible
    \item Ensure citations match the bibliography
    \item Proofread citations carefully
\end{itemize}

\section{Conclusion}
Proper citation is essential for academic writing. It demonstrates your knowledge of the field, gives credit to other researchers, and allows readers to verify your claims.

% Bibliography
\printbibliography

\end{document}
